# Analysis Script for the Systematic Review

## Project: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'

This script closely follows the preregistration, which is uploaded on OSF.

```{r setup}
# Packages used for the data workflow

library(readxl)
library(rentrez)
library(XML)
library(purrr)
library(ggplot2)
library(MetBrewer)
```

#### Get search data

The search was conducted on the 16th March 2022 at around 10 pm CET.

The Pubmed results are saved in the file 'pubmed.csv'; the Web of Science results are saved in the file 'wos.xls'.

```{r read, warning=FALSE, message=FALSE}
# file name for PubMed results
pm_file <- "../data/search/pubmed.csv"

# file name for merged WoS results
wos_file <- "../data/search/wos.xls"

# read PubMed data
pm_data_pre <- read.csv(pm_file)
pm_data <- data.frame(
  pmid = pm_data_pre[, 1],
  title = pm_data_pre$Title,
  authors = pm_data_pre$Authors,
  journal = pm_data_pre$Journal.Book,
  year = pm_data_pre$Publication.Year,
  doi = pm_data_pre$DOI,
  abstract = NA
)
# read Web of Science data
wos_data_pre <- readxl::read_xls(wos_file)
wos_data <- data.frame(
  pmid = wos_data_pre$`Pubmed Id`,
  title = wos_data_pre$`Article Title`,
  authors = wos_data_pre$`Author Full Names`,
  journal = wos_data_pre$`Source Title`,
  year = wos_data_pre$`Publication Year`,
  doi = wos_data_pre$DOI,
  abstract = wos_data_pre$Abstract
)

# count PubMed results
nrow(pm_data)
# count Web of Science results
nrow(wos_data)
```

#### Removal of duplicates

```{r duplicates}
# Merge data frame
merge_data <- rbind(wos_data, pm_data)

# Filter entries without doi
no_doi <- is.na(merge_data$doi) | (merge_data$doi == "")
merge_data <- merge_data[!no_doi, ]

# Count removal from missing doi
sum(no_doi)

# Find duplicated DOIs
dupl <- duplicated(merge_data$doi)
# Filter duplicates
merge_data <- merge_data[!dupl, ]

# Count removal of duplicates
sum(dupl)
```

#### Automated title filtering

```{r filter}
# Filter titles by terms
terms <- paste0("review|comment|correction|retraction|meta-analysis",
                "|editorial|erratum|reply")
excl <- grepl(terms, merge_data$title, ignore.case = TRUE)
filtered_data <- merge_data[!excl, ]

# Count removal by title filtering
sum(excl)
# count remaining articles
nrow(filtered_data)
```

#### Random Sampling

```{r sampling, eval=FALSE}
# Assign ids to articles left
filtered_data$id <- seq_len(nrow(filtered_data))

# randomly sample 500 articles
set.seed(4711)
smpl <- sample(filtered_data$id, size = 500)

sampled_data <- filtered_data[smpl, ]
# assign sample id
sampled_data$sid <- seq_len(nrow(sampled_data))

# get abstract from PubMed where missing

no_abstract <- is.na(sampled_data$abstract)

no_abstract_pmid <- sampled_data$pmid[no_abstract]

# function to scrape abstract from the PubMed API
get_abstract <- function(pmid) {
  if (is.na(pmid))
    return(NA)
  xml_data <- rentrez::entrez_fetch(
    db = "pubmed", 
    id = pmid, 
    rettype = "xml", 
    parse = TRUE
  )
  text <- XML::xmlValue(XML::getNodeSet(xml_data, "//AbstractText"))
  out <- paste(text, collapse = "///")
  if (out == "") out <- NA
  out
}

# retrieve missing abstracts
sampled_data$abstract[no_abstract] <- purrr::map_chr(
  .x = no_abstract_pmid, 
  .f = get_abstract
)

# manually add abstract for articles, where automated abstract retrieval did not
# work. In these cases no blinding applies to the principal investigator
which(is.na(sampled_data$abstract))

sampled_data[500,]

sampled_data$abstract[50] <- readtext::readtext("../data/search/a50.txt")$text
sampled_data$abstract[238] <- readtext::readtext("../data/search/a238.txt")$text
sampled_data$abstract[288] <- readtext::readtext("../data/search/a288.txt")$text
sampled_data$abstract[416] <- readtext::readtext("../data/search/a416.txt")$text
sampled_data$abstract[488] <- readtext::readtext("../data/search/a488.txt")$text
sampled_data$abstract[490] <- readtext::readtext("../data/search/a490.txt")$text
sampled_data$abstract[500] <- readtext::readtext("../data/search/a500.txt")$text

# later plotting throws an error for some articles due to unrecognized html-tags
# in the abstracts. These abstract are replaced by their text equivalents
sampled_data$abstract[344] <- readtext::readtext("../data/search/a344.txt")$text
sampled_data$abstract[356] <- readtext::readtext("../data/search/a356.txt")$text

save(sampled_data, file = "../data/screen/sample.Rda")
```

#### Prepare Title-Abstract plots for screening

```{r abstractplots, eval=FALSE}
load("../data/screen/sample.Rda")

# Function for creating plots with blinded abstract data
create_abstract_plot <- function(row, data) {
  # get random colour
  # different colours in plots are used to stay concentrated when scanning
  # the abstracts
  clrs <- c(MetBrewer::met.brewer("Moreau")[-4])
  clr <- sample(clrs, 1)
  
  g <- ggplot(data = NULL, aes(x = 0, y = 1)) +
  ggtext::geom_textbox(
    width = 1, 
    box.colour = "transparent", 
    label =  gsub(">", "", data$abstract[row]),
    colour = clr
  ) +
  labs(
    title = paste0(strwrap(data$title[row], 80), collapse = "\n"), 
    subtitle = paste0("Sampling ID: ", sprintf("%03i", data$sid[row])),
    caption = paste0("ID: ", data$id[row], "; PMID = ", data$pmid[row])
    ) +
  theme_void()
  
  ggsave(
    filename = paste0("../plots/abstracts/", sprintf("%03i", data$sid[row]), ".png"), 
    plot = g, 
    dpi = 300, width = 7, height = 5, bg = "white"
  )
}


# Create blinded abstract plots
purrr::walk(
  .x = seq_len(nrow(sampled_data)),
  .f = create_abstract_plot, 
  data = sampled_data
)

```

