---
title: " "
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: TRUE
mainfont: Arial
fontsize: 11pt
urlcolor: blue
linkcolor: black
geometry: "left = 4cm, right = 2cm, top = 2cm, bottom = 2cm"
header-include:
  - \usepackage{setspace}\spacing{1.5}
  - \renewcommand{\contentsname}{Inhalt}
bibliography: ../references.bib
csl: '`r here::here("apa.csl")`'
params:
  title: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'
  name: Simon Nolte
  university: German Sport University Cologne
  year: 2022
---

```{r setup, include=FALSE}

```

\begin{centering}

\vspace*{2cm}

\LARGE
{\bf `r params$title`}

\vspace{5cm}

\LARGE
{\bf Bachelor thesis}

\LARGE
{\bf from}

\vspace{2cm}

\LARGE
{\bf `r params$name`}

\vspace{6cm}

\LARGE
{\bf `r params$university`}

\Large
{\bf `r paste0("Cologne ", params$year)`}

\end{centering}

\pagenumbering{gobble}

\newpage

Thesis supervisor:

Dr. Oliver Jan Quittmann

Institute of Movement and Neurosciences

\newpage

Affirmation in lieu of an oath 
 
Herewith I affirm in lieu of an oath that I have authored this Master thesis independently and did not use any other sources and tools than indicated. All citations, either direct quotations or passages which were reproduced verbatim or nearby-verbatim from publications, are indicated and the respective references are named. The same is true for tables and figures. I did not submit this piece of work in the same or similar way or in extracts in another assignment.


\_____________________________

Personally signed

\newpage

# Zusammenfassung (German abstract) {-}

\newpage

\pagenumbering{roman}

# Table of contents {-}

\renewcommand{\contentsname}{}

\tableofcontents

\newpage

\renewcommand{\listfigurename}{}

# List of Figures {-}

\listoffigures

\newpage

\renewcommand{\listtablename}{}

# List of Tables {-}

\listoftables

\newpage

\pagenumbering{arabic}

```{r intro, child = 'introduction.Rmd'}

```

\newpage

```{r methods, child = 'methods.Rmd'}

```

\newpage

```{r results, child = 'results.Rmd'}

```

\newpage

```{r discussion, child = 'discussion.Rmd'}

```

\newpage

```{r conclusion, child = 'conclusion.Rmd'}

```

\newpage

# Bibliography

<div id="refs"></div>

\newpage

# Appendix
